# Install R & RStudio {#install-r-and-rstudio}

## R

```{r installr, echo=FALSE}
knitr::include_url("https://www.youtube-nocookie.com/embed/VJ6-ZxVUNzw")
```

If you prefer written instructions over video, follow these steps:

- Download R for your operating system from a [CRAN mirror](https://cran.r-project.org/mirrors.html)
  (Windows, macOS, or Linux installers are listed there).
- Run the installer with default options.
- Verify the installation by opening a terminal and running `R.version.string`.

## RStudio

```{r installrstudio, echo=FALSE}
knitr::include_url("https://www.youtube-nocookie.com/embed/_askhjFWlHA")
```

If you prefer written instructions over video, follow these steps:

- Download RStudio Desktop from [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/).
- Install it after installing R (RStudio needs R to work).
- Open RStudio and create a new project via *File > New Project* (this creates
  an `.Rproj` file that keeps your working directory, history, and settings together).
- Verify your setup by running `R.version.string` in the R console.