# Dataframes {#dataframes}

## Introduction

In this chapter, we will learn to:

- create dataframe
- select columns 
- select rows
- utitlity functions

## Create dataframes

Use `data.frame` to create dataframes. Below is the function syntax:

```{r df1}
args(data.frame)
```

Data frames are basically lists with elements of equal lenght and as such, they are heterogeneous. Let us create a dataframe:

```{r df2}
name <- c('John', 'Jack', 'Jill')
age <- c(29, 25, 27)
graduate <- c(TRUE, TRUE, FALSE)
students <- data.frame(name, age, graduate)
students
```

## Basic Information

```{r df3, collapse=TRUE}
class(students)
names(students)
colnames(students)
str(students)
dim(students)
nrow(students)
ncol(students)
```

## Select Columns

### Single Column

- `[]`
- `[[]]`
- `$`

```{r df4, collapse=TRUE}
# using [
students[1]

# using [[
students[[1]]

# using $
students$name
```

### Multiple Columns

```{r df5, collapse=TRUE}
students[, 1:3]

students[, c(1, 3)]
```

## Select Rows

```{r df6, collapse=TRUE}
# single row
students[1, ]

# multiple row
students[c(1, 3), ]
```

If you have observed carefully, the column `names` has been coerced to type 
factor. This happens because of a default argument in `data.frame` which is 
`stringsAsFactors` which is set to `TRUE`. If you do not want to treat it as 
`factors`, set the argument to `FALSE`.

```{r df7}
students <- data.frame(name, age, graduate, stringsAsFactors = FALSE)
```



