# Dataframes in R {#data-frames-in-r}

## Introduction

In this chapter, we will learn to:

- create dataframe
- select columns 
- select rows
- utility functions

## Create dataframes

Use `data.frame` to create dataframes. Below is the function syntax:

```{r df1}
args(data.frame)
```

Data frames are basically lists with elements of equal length and as such, they are heterogeneous. Let us create a dataframe:

```{r df2}
name <- c('John', 'Jack', 'Jill')
age <- c(29, 25, 27)
graduate <- c(TRUE, TRUE, FALSE)
students <- data.frame(name, age, graduate)
students
```

## Basic Information

```{r df3, collapse=TRUE}
class(students)
names(students)
colnames(students)
str(students)
dim(students)
nrow(students)
ncol(students)
```

## Select Columns

### Single Column

- `[]`
- `[[]]`
- `$`

```{r df4, collapse=TRUE}
# using [
students[1]

# using [[
students[[1]]

# using $
students$name
```

### Multiple Columns

```{r df5, collapse=TRUE}
students[, 1:3]

students[, c(1, 3)]
```

## Select Rows

```{r df6, collapse=TRUE}
# single row
students[1, ]

# multiple row
students[c(1, 3), ]
```

Starting with R 4.0.0 (April 2020), the default value of `stringsAsFactors`
in `data.frame()` is `FALSE`, meaning character columns remain as character
type. In older versions of R (< 4.0.0), the default was `TRUE`, which
automatically converted character vectors to factors.

```{r df7}
students <- data.frame(name, age, graduate, stringsAsFactors = FALSE)
```



