# R Packages {#packages}

## Introduction

In this chapter, we will learn about R packages. Packages are fundamental to R. There are approximately 15000 packages available on [CRAN](https://cran.r-project.org/) or the Comprehensive R Archive Network.

Packages are available for different [topics](https://cran.r-project.org/web/views/). You should always look for a package before writing code from scratch. In case you have written your own codes for a new analysis or topic, do share it with the R community by converting the code into a package. You can learn more about building R packages from [R Packages](https://r-pkgs.had.co.nz/), a book written by [Hadley Wickham](https://hadley.nz/).

In this post, we will learn to:

- install R packages from 
    - CRAN
    - GitHub
    - BitBucket
    - Bioconductor
    - rForge
- install different versions of a package
- load, update & remove installed packages
- access package documentation

## Install Packages

### CRAN

Packages from CRAN can be installed using `install.packages()`. The name of the package must be enclosed in single or double quotes.

```{r install1, eval=FALSE}
install.packages('ggplot2')
```

### GitHub

Some R packages are made available on [GitHub](https://github.com/) before releasing them on CRAN. Such packages can be installed using `install_github()` from [devtools](https://cran.r-project.org/web/packages/devtools/index.html) or [remotes](https://cran.r-project.org/web/packages/remotes/index.html) package. You need tp specify the name of the repository and the package. For example, to download [ggplot2](https://ggplot2.tidyverse.org/) or [dplyr](https://dplyr.tidyverse.org/), below is the code:

```{r install2, eval=FALSE}
devtools::install_github("tidyverse/ggplot2")
remotes::install_github("tidyverse/dplyr")
```

### BitBucket

[Bitbucket](https://bitbucket.org/) is similar to GitHub. You can install packages from Bitbucket using `install_bitbucket()` from devtools or remotes pacakge.

```{r install3, eval=FALSE}
devtools::install_bitbucket("dannavarro/lsr-package")
remotes::install_bitbucket("dannavarro/lsr-package")
```

### Bioconductor

[Bioconductor](https://www.bioconductor.org/) provides tools for analysis and comprehension of high throughput genomic data. Packages hosted on Bioconductor can be installed in multiple ways:

#### devtools

Use `install_bioc()` from devtools.

```{r install4, eval=FALSE}
install_bioc("SummarizedExperiment")
```

#### biocLite

Use `biocLite()` function.

```{r install5, eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite('GenomicFeatures')
```

### rForge

Many R packages are hosted at [R-Forge](https://r-forge.r-project.org/), a platform for development of R packages. 

```{r install6, eval=FALSE}
install.packages('quantstrat', repos = 'https://r-forge.r-project.org/')
```

## Install Different Versions

Now that we have learnt how to install packages, let us look at installing different versions of the same package.

```{r install7, eval=FALSE}
remotes::install_version('dplyr', version = 0.5.0)
```

If you want to install the latest release from GitHub, append `@*release` to the repository name. For example, to install the latest release of dplyr:

```{r install8, eval=FALSE}
remotes::install_github('tidyverse/dplyr@*release')
```

## Installed Packages

- `installed.packages()`: view currently installed packages
- `library('package_name')`: load packages
- `autoload('function_name', 'package_name')`: load functions and data from packages only when called in the script
- `available.package()`: packages available for installation
- `old.packages()`: packages which have new versions available
- `new.packages()`: packages already not installed
- `update.packages()`: update packages which have new versions
- `remove.packages('package_name')`: remove installed packages

## Library Paths

Library is a directory that contains all installed packages. Usually there will be more than one R library in your systme. You can find the location of the libraries using `.libPaths()`.

```{r install9}
.libPaths()
```

You can use `lib.loc` when you want to install, load, update and remove packages from a particular library.

### Install

```{r install10, eval=FALSE}
install.packages('stringr', lib.loc = "C:/Program Files/R/R-3.4.1/library")
```

### Load

```{r install11, eval=FALSE}
library(lubridate, lib.loc = "C:/Program Files/R/R-3.4.1/library")
```

### Update Packages

```{r install12, eval=FALSE}
update.packages(lib.loc = "C:/Program Files/R/R-3.4.1/library")
```

### Remove Packages

```{r install13, eval=FALSE}
remove.packages(lib.loc = "C:/Program Files/R/R-3.4.1/library")
```

