# Dataframes in R {#sec-data-frames-in-r}

> **Learning Objectives:** After this chapter you will be able to:
> - Create a data frame from equal-length vectors.
> - Inspect one with `str()`, `dim()`, `names()`, and `nrow()`/`ncol()`.
> - Select columns with `[`, `[[`, and `$`.
> - Select rows by position and by condition.
> - Subset rows and columns independently.
> - Explain why `stringsAsFactors` defaults to `FALSE` in R 4.0.0 and later.

## Introduction

Data frames are lists of equal-length vectors, so they can hold columns of
different types.

## Create dataframes

Use `data.frame` to create dataframes. Below is the function syntax:

```{r df1}
args(data.frame)
```

Data frames are basically lists with elements of equal length and as such, they are heterogeneous. Let us create a dataframe:

```{r df2}
name <- c('John', 'Jack', 'Jill')
age <- c(29, 25, 27)
graduate <- c(TRUE, TRUE, FALSE)
students <- data.frame(name, age, graduate)
students
```

## Basic Information

```{r df3, collapse=TRUE}
class(students)
names(students)
colnames(students)
str(students)
dim(students)
nrow(students)
ncol(students)
```

## Select Columns

### Single Column

- `[]`
- `[[]]`
- `$`

```{r df4, collapse=TRUE}
# using [
students[1]

# using [[
students[[1]]

# using $
students$name
```

### Multiple Columns

```{r df5, collapse=TRUE}
students[, 1:3]

students[, c(1, 3)]
```

## Select Rows

```{r df6, collapse=TRUE}
# single row
students[1, ]

# multiple row
students[c(1, 3), ]
```

## Try it in your browser {.unnumbered}

Edit the code below and press **Run**. It executes entirely in your browser via
[WebR](https://docs.r-wasm.org/webr/latest/) — no R installation needed. The
runtime downloads once, on this page only.

```{webr-r}
students <- data.frame(
  name = c("Asha", "Ravi", "Meera", "Arjun"),
  age = c(21, 22, 20, 23),
  graduate = c(TRUE, TRUE, FALSE, TRUE)
)

# rows and columns independently
students[2, ]
students[, "age"]

# filter by condition
students[students$age > 21, ]

# how many rows and columns
dim(students)
```

Starting with R 4.0.0 (April 2020), the default value of `stringsAsFactors`
in `data.frame()` is `FALSE`, meaning character columns remain as character
type. In older versions of R (< 4.0.0), the default was `TRUE`, which
automatically converted character vectors to factors.

> **Common Mistake:** Assuming `data.frame()` still converts strings to factors.
> That was the default before R 4.0.0 and was a frequent source of confusion.
> On R >= 4.0.0 the default is `FALSE`.

```{r df7}
students <- data.frame(name, age, graduate, stringsAsFactors = FALSE)
```

## Practice Check

**Knowledge Check:** *One column, three operators*
What do `students[1]`, `students[[1]]`, and `students$name` each return?

<details>
<summary>Click to reveal answer</summary>

**Answer:** A one-column **data frame**; a **vector**; and the same **vector**.

*Explanation:* `[` always returns a data frame, even for a single column, so
downstream code that expects a vector breaks. `[[` and `$` both extract the
column itself as a vector. Reach for `[[` or `$` whenever you want the values.
</details>

**Knowledge Check:** *Rows and columns are indexed separately*
What does `students[1, ]` return, and how does that differ from
`students[, 1]`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `students[1, ]` returns the **first row**; `students[, 1]` returns
the **first column**. The comma is what separates the two.

*Explanation:* Inside `[` the first position selects rows and the second selects
columns. Leaving one side empty means "all of them". So `students[1, ]` is
row 1, `students[, 1]` is column 1, and `students[1, 1]` is the single cell
where they cross.
</details>

**Knowledge Check:** *Filtering rows by a condition*
How do you return only the students older than 26?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `students[students$age > 26, ]`, which gives John (29) and Jill (27).

*Explanation:* The condition builds a logical vector with one value per row.
Using it as the row index keeps the rows where the condition is `TRUE` and drops
the rest. Because the condition has the same length as the data frame, the
column is left out of the result.
</details>

**Knowledge Check:** *How many rows and columns?*
The data frame has three columns and three rows. Which two functions report
those counts directly?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `nrow(students)` gives `3` and `ncol(students)` gives `3`. Both are
also available together as `dim(students)`, which returns `c(3, 3)`.

*Explanation:* `dim()` returns rows first, then columns, so `dim(students)[1]`
is the row count. `str()` gives you a structural overview while `nrow()` and
`ncol()` answer the two specific questions directly.
</details>

**Knowledge Check:** *`stringsAsFactors` changed in R 4.0.0*
In R 4.0.0 and later, what does `data.frame(name, age, graduate)` do with the
character column `name`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** It keeps it as `character`. Before R 4.0.0 the default was `TRUE`,
which silently converted character columns to factors.

*Explanation:* The old default caused endless confusion, because strings became
categories and levels were alphabetised behind your back. Passing
`stringsAsFactors = FALSE` is now redundant on R >= 4.0.0, but it keeps code
readable and portable to older R.
</details>



