# Lists in R {#sec-lists-in-r}

> **Learning Objectives:** After this chapter you will be able to:
> - Create a list that holds elements of different types.
> - Name list elements with `names()`.
> - Access elements with `[[`, `[`, and `$`.
> - Explain why `[[` returns the element while `[` returns a list.
> - Convert a vector or matrix into a list with `as.list()`.
> - Convert a list back into a vector with `unlist()`.

## Introduction

Lists are very useful as they are heterogeneous i.e. they can contain different data types. If you remember, vectors and matrices are homogeneous i.e. they can contain only one type of data. If you include different data types, they will all be coerced to the same type. With lists, it is different. In this chapter, we will delve deeper into lists.

## Creating Lists

To create a list, we use the `list()` function. Let us create a simple list to demonstrate how they can contain different data types.

```{r list1}
# numeric vector
vect1 <- 1:10

# character vector 
vect2 <- c('Jack', 'John', 'Jill')

# logical vector
vect3 <- c(TRUE, FALSE)

# matrix
mat <- matrix(data = 1:9, nrow = 3)

# list
l <- list(vect1, vect2, vect3, mat)
l
```

If you observe the output, all the elements of the list retain their data types. Now let us learn how to access the elements of the list.

## Access List Elements

You can access the elements of a list using the following operators:

- `[[`
- `[`
- `$`

Let us try them one by one.

```{r list2}
# using [[
l[[1]]

# using [
l[1]
```

`[[` extracts the element itself (which may be a vector, matrix, data frame, or nested list), while `[` returns a new list containing the selected element(s). The `$` operator can be used only when we have named elements in the list. Let us add names to the elements. Use the `names()` function to add names to the list.

```{r list3}
# named elements
names(l) <- c('vect1', 'vect2', 'vect3', 'mat')
l

# use $
l$vect1

# use [[
l[['vect1']]

# use [
l['vect1']
```

## Coerce other objects

You can coerce other objects to list using `as.list()`.

```{r list4}
# vector
vect1 <- 1:10
as.list(vect1)

# matrix
mat <- matrix(data = 1:9, nrow = 3)
as.list(mat)
```

## Coerce list to other objects

Use `unlist()` to coerce a list to a vector.

```{r list5}
# unlist
unlist(l)

# remove names
unlist(l, use.names = FALSE)
```

## Practice Check

**Knowledge Check:** *Why use a list at all?*
`vect1 <- 1:10`, `vect2 <- c('Jack', 'John', 'Jill')`, and `vect3 <- c(TRUE, FALSE)`.
Why not just `c(vect1, vect2, vect3)`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** That would coerce everything to `character`, because `character` is
the most flexible type. A list keeps each element's own type.

*Explanation:* `c(vect1, vect2, vect3)` gives you
`c("1", "2", ..., "10", "Jack", ...)`. Vectors and matrices are homogeneous, so
mixing types destroys information. `list(vect1, vect2, vect3)` stores the three
vectors side by side with their types intact.
</details>

**Knowledge Check:** *`[[` versus `[`*
Given `l <- list(1:10, c('Jack', 'John'))`, what is the difference between
`l[[1]]` and `l[1]`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `l[[1]]` is the vector `1:10` itself. `l[1]` is a list containing
one element, which is that same vector.

*Explanation:* `[[` extracts the element, so you get the object and can operate
on it directly. `[` builds a new list holding the selection, which is what you
want when you want to delete or keep several elements. Square brackets are for
selecting; double square brackets are for extracting.
</details>

**Knowledge Check:** *`$` needs names*
Why does `l$vect1` fail before you run `names(l) <- c('vect1', 'vect2', 'vect3', 'mat')`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** The elements have no names, so there is nothing for `$` to match.

*Explanation:* `$` looks up elements **by name**, not by position. Until the
list has names, use `l[[1]]` or `l[1]`. Once named, `l$vect1` and
`l[['vect1']]` both return the vector.
</details>

**Knowledge Check:** *Coercing to a list*
How do you turn the vector `vect1 <- 1:10` into a list, one element per value?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `as.list(vect1)`, which gives a list of ten single numbers.

*Explanation:* `as.list()` splits the vector into its individual elements. The
same call works on a matrix, in which case it splits the matrix into columns.
The reverse direction is `unlist()`.
</details>

**Knowledge Check:** *Flattening a list*
What does `unlist(l, use.names = FALSE)` do differently from `unlist(l)`?

<details>
<summary>Click to reveal answer</summary>

**Answer:** It returns the same values but without the names.

*Explanation:* `unlist()` concatenates every element into one vector and keeps
the names it finds, so you get a named vector. Setting `use.names = FALSE` drops
the names, which is usually what you want for further calculation.
</details>
