# RStudio Projects & Data Import {#sec-projects-import}

> **Learning Objectives:** After this chapter you will be able to:
>
> - Work inside an RStudio Project (`.Rproj`) instead of `setwd()`.
> - Build portable paths with `here::here()`.
> - Read CSV files with `readr::read_csv()` and Excel files with `readxl::read_excel()`.

```{r proj-setup, message=FALSE}
library(readr)
library(readxl)
library(here)
```

## Why avoid `setwd()`

`setwd("C:/Users/you/project")` hard-codes a path that exists only on your
machine. Anyone else who opens the script — or you on a different computer —
gets an error. An RStudio Project (`.Rproj`) fixes this: opening the `.Rproj`
file sets the working directory to the project folder automatically, so
relative paths just work.

**Workflow:** File → New Project → New Directory (or Existing Directory) →
give it a name. From then on, open the `.Rproj` file, not individual scripts.

## Portable paths with `here`

The `here` package builds paths from the project root, wherever the project
lives on disk:

```{r here-demo}
here("data", "students.csv")
```

> **Common Mistake:** Paths with `../` or `C:/...` break when the project
> moves. If a path starts with a drive letter or `../`, rewrite it with
> `here()`.

## Reading CSV files

Practice data lives in the `data/` directory (CC0 / public domain, created
for this book). Read it with `read_csv()`:

```{r read-csv}
students <- read_csv(here("data", "students.csv"))
students
```

## Reading Excel files

`read_excel()` works the same way for `.xlsx` files — sheet 1 by default:

```{r read-excel}
students_xl <- read_excel(here("data", "students.xlsx"))
students_xl
```

## Try it in your browser {.unnumbered}

Edit the code below and press **Run**. It executes entirely in your browser via
[WebR](https://docs.r-wasm.org/webr/latest/) — no R installation needed. The
runtime downloads once, on this page only.

```{webr-r}
library(readr)

# the browser filesystem starts empty, so fetch the sample first
if (!file.exists("students.csv")) {
  download.file(
    "https://intro-r.rsquaredacademy.com/data-webr/students.csv",
    "students.csv"
  )
}

students <- read_csv("students.csv", show_col_types = FALSE)
students

# column types are parsed, not guessed at read time
sapply(students, class)
```

## Practice Check

**Knowledge Check:** *Project-relative paths*
Why does `read_csv(here("data", "students.csv"))` keep working when you move
the whole project folder to another computer, but
`read_csv("C:/Users/you/project/data/students.csv")` does not?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `here()` resolves the path relative to the project root at run
time, so it adapts to the new location. The `C:/...` path is absolute and
points to a folder that does not exist on the other computer.

*Explanation:* Projects + `here()` keep every path portable
(`project root` → `data` → `file`); absolute paths and `setwd()` tie the
script to one machine.
</details>

**Knowledge Check:** *Why `.Rproj` beats `setwd()`*
What breaks when you share a script containing
`setwd("C:/Users/you/project")` with a colleague?

<details>
<summary>Click to reveal answer</summary>

**Answer:** The script errors because that folder does not exist on your
colleague's machine.

*Explanation:* `setwd()` with an absolute path ties the script to one
computer; opening a shared `.Rproj` file sets everyone's working directory
to the project folder automatically.
</details>

**Knowledge Check:** *`here()` resolution*
Your project lives at `D:/work/survey` and contains `data/resp.csv`.
What does `here("data", "resp.csv")` return?

<details>
<summary>Click to reveal answer</summary>

**Answer:** `D:/work/survey/data/resp.csv` (i.e. `project root` → `data` →
`resp.csv`).

*Explanation:* `here()` anchors paths at the project root (where the
`.Rproj` file lives), joining each additional component with the correct
separator for your OS.
</details>

**Knowledge Check:** *`read_csv()` vs `read.csv()`*
Name one practical advantage of `readr::read_csv()` over base R's
`read.csv()` for a beginner.

<details>
<summary>Click to reveal answer</summary>

**Answer:** Any one of: it never converts strings to factors, it prints a
compact tibble preview, or it reports column types as it reads.

*Explanation:* `read_csv()` uses modern defaults (no `stringsAsFactors`
surprises) and tells you the parsed type of each column, so mismatches are
visible immediately.
</details>

**Knowledge Check:** *Excel sheets*
`read_excel(here("data", "students.xlsx"))` reads which sheet by default,
and how would you read the second sheet instead?

<details>
<summary>Click to reveal answer</summary>

**Answer:** The first sheet by default; pass `sheet = 2` (or the sheet name)
to read the second sheet.

*Explanation:* `read_excel()` defaults to sheet 1; the `sheet` argument
selects another sheet by position or name.
</details>
