# Your First Visualization {#first-plot}

> **Learning Objectives:** After this chapter you will be able to:
> - Load `ggplot2` and its bundled `mpg` dataset.
> - Draw the same scatterplot in base R and in ggplot2.
> - Describe a plot as data + aesthetics + geometry.

```{r plot-setup, message=FALSE}
library(ggplot2)
# ggplot2 bundles the `mpg` dataset (GPL-2 | GPL-3 license)
data(mpg, package = "ggplot2")
```

> **Common Mistake:** Base R has no `mpg` dataset (it has `mtcars`).
> `plot(mpg$displ, mpg$hwy)` without `library(ggplot2)` fails with
> `object 'mpg' not found`. Always load `ggplot2` first.

## Base R graphics

```{r base-plot}
plot(mpg$displ, mpg$hwy, xlab = "Displacement", ylab = "Highway MPG")
```

## ggplot2 graphics

```{r ggplot}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  labs(x = "Displacement", y = "Highway MPG")
```

**Mental model:** every ggplot is *data* (`mpg`) + *aesthetics* (`aes(x, y)`)
+ *geometry* (`geom_point()`). Change the geometry, keep the mapping.

## Practice Check

**Knowledge Check:** *Data + aesthetics + geometry*
In `ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()`, which part is the
data, which is the aesthetics mapping, and which is the geometry?

<details>
<summary>Click to reveal answer</summary>

**Answer:** Data: `mpg`. Aesthetics: `aes(x = displ, y = hwy)`.
Geometry: `geom_point()`.

*Explanation:* `ggplot()` takes the data and the variable mapping;
each `geom_*()` layer decides how those mapped variables are drawn.
</details>
