# Your First Data Wrangling {#first-wrangle}

> **Learning Objectives:** After this chapter you will be able to:
> - Use the five core `dplyr` verbs: `filter()`, `select()`, `mutate()`,
>   `arrange()`, `summarise()`.
> - Chain steps with the native base pipe `|>`.
> - Know where to go next after this book.

```{r wrangle-setup, message=FALSE}
library(dplyr)
# ggplot2 bundles the `mpg` dataset (GPL-2 | GPL-3 license)
data(mpg, package = "ggplot2")
```

We use the native base pipe `|>` (standard since R 4.1.0; zero external
dependencies). We do **not** use `%>%`.

## The five verbs

```{r verbs}
mpg |>
  filter(manufacturer == "audi") |>
  select(manufacturer, model, displ, hwy) |>
  mutate(km_per_litre = hwy * 0.425144) |>
  arrange(desc(hwy))
```

```{r summarise}
mpg |>
  summarise(mean_hwy = mean(hwy), n = dplyr::n())
```

> **Common Mistake:** `mean()` on a vector with `NA` returns `NA`.
> Add `na.rm = TRUE` when missing values are present.

## Practice Check

**Knowledge Check:** *Filter then select*
What does `mpg |> filter(manufacturer == "audi") |> select(model, hwy)` return?

<details>
<summary>Click to reveal answer</summary>

**Answer:** A data frame with two columns (`model`, `hwy`) containing only
the rows where `manufacturer` is `"audi"`.

*Explanation:* `filter()` keeps matching rows; `select()` keeps matching
columns. The `|>` pipe passes the left-hand result into the first argument
of the right-hand function.
</details>

## Where to go next

> **Where to go next:**
> - Book: [R for Data Science (2e)](https://r4ds.hadley.nz/) — start at Chapter 1
> - Courses: [Rsquared Academy](https://www.rsquaredacademy.com/) Data Wrangling series
> - More books: [ebooks.rsquaredacademy.com](https://ebooks.rsquaredacademy.com/)
